---
title: "Analysis of organelle segmentations"
author: "Daniel J Kelpsch"
format: html
editor: visual
---

Working on analysis of antibody panel 1 for the initial (automated and manual) plates set up in iNDI production. Nucleus segmentation identified \~2M nuclei (and corresponding ROIs that equate to cell bodies). Using that ROI information, segmentation for lysosomes, endosomes, and mitochondria was preformed on each respective channel and initial features were extracted.

Here, the goal is to import all organelle and nucleus data, preform all cleanup, and output final dataframes for downstream analysis.

## R set up

```{r}
#| label: Set project working directory
setwd("/vf/users/kelpschdj/iNDI/Production/Organelle_segmentation/Panel_1/analysis")
```

```{r}
#| include: false
#| label: Load required ackages

package_list <- c("tidyverse", "uwot", "ranger", "arrow", "glue", "corrplot")
new_packages <- package_list[!(package_list %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)
lapply(package_list, library, character.only = TRUE)
```

## File import

```{r}
#| label: Import essential data
# samplesheets
samplesheet_auto_fn <- "/vf/users/CARDPB2/iNDI/Production/metadata/all_automated_plates_combined.csv"
samplesheet_auto <- read_csv(samplesheet_auto_fn) %>%
  mutate(
    well_row = match(str_extract(Well_ID, "^[A-Z]"), LETTERS),
    well_column = as.integer(str_extract(Well_ID, "\\d+")),
    experiment_design = "Automated"
  )

samplesheet_man_fn <- "/vf/users/CARDPB2/iNDI/Production/metadata/all_manual_plates_combined.csv"
samplesheet_man <- read_csv(samplesheet_man_fn) %>%
  mutate(
    well_row = match(str_extract(Well_ID, "^[A-Z]"), LETTERS),
    well_column = as.integer(str_extract(Well_ID, "\\d+")),
    experiment_design = "Manual"
  )

samplesheet <- samplesheet_auto %>% 
  add_row(samplesheet_man)

# plate ID converter
plateID_to_folder_ID_fn <- "/vf/users/CARDPB2/iNDI/Production/metadata/indi_plateID_to_folderID.csv"
plateID_to_folder_ID <- read_csv(plateID_to_folder_ID_fn)

# organelle segmentations
df <- read_parquet("/vf/users/kelpschdj/iNDI/Production/Organelle_segmentation/Panel_1/organelle_segmentation_AbPanel1.parquet")

df <- df %>%
  mutate(
    well_row    = as.integer(str_extract(channel_filename, "(?<=r)\\d+")),
    well_column = as.integer(str_extract(channel_filename, "(?<=c)\\d+"))
  ) %>%
  left_join(plateID_to_folder_ID, by = join_by(Measurement_ID == Folder_ID)) %>%
  left_join(samplesheet, by = c("well_row", "well_column", "iNDI_Plate_ID")) %>%
  mutate(Genotype_Name = str_replace_all(Genotype_Name, "FUS R459X", "FUS_R495X")) # Genotype_Name has FUS R459X* when it is supposed to be FUS_R495X*, manual fix 
```

### Data cleaning

```{r}
# Fix genotype flags (for parental)
# Genotype column is a bit weird, WT is just NA, just need to put something in there (WT/WT probably)
df %>% pull(Genotype) %>% factor() %>% levels()

# Gah all the manual plates are not annotated the same... 
df %>% filter(is.na(Genotype)) %>% pull(Genotype_Name) %>% factor() %>% levels()

# What is REV/Y?
df %>% filter(Genotype == "REV/Y") %>% select(Genotype, Genotype_Name) %>% pull(Genotype_Name) %>% factor() %>% levels()

# Just change Genotype to "WT/WT" when Genotype_name is "Parental"
df <- df %>%
  mutate(
    Genotype = case_when(
      Genotype_Name == "Parental" ~ "WT/WT",
      .default = Genotype
      )
    )
```

### Import nucleus segmentation data

```{r}
#| label: Extract nucleus data

nuc_files <- list.files(
  path       = "/vf/users/kelpschdj/iNDI/Production/Nucleus_segmentation/Nucleus_feature_data/",
  pattern    = "\\.csv$",
  full.names = TRUE
)

px_factor <- 0.09418670872212731
radius    <- 120
img_size  <- 2160

flag_size <- function(plate, min_um2 = 20, max_um2 = 200) {
  plate %>%
    mutate(
      area_um2  = area * px_factor^2,
      fail_size = area_um2 <= min_um2 | area_um2 >= max_um2
    )
}

flag_edge <- function(plate, radius = 120, img_h = img_size, img_w = img_size) {
  plate %>%
    mutate(
      fail_edge = `centroid-0` - radius < 0  |
                  `centroid-0` + radius > img_h |
                  `centroid-1` - radius < 0  |
                  `centroid-1` + radius > img_w
    )
}

flag_overlap_img <- function(df, radius = 120, max_overlap_frac = 0.05) {
  circle_area <- pi * radius^2
  max_overlap <- max_overlap_frac * circle_area

  cx   <- df$`centroid-1`
  cy   <- df$`centroid-0`
  n    <- nrow(df)
  fail <- logical(n)

  for (i in seq_len(n)) {
    dx          <- cx - cx[i]
    dy          <- cy - cy[i]
    d           <- sqrt(dx^2 + dy^2)
    neighbours  <- which(d > 0 & d < 2 * radius)

    if (length(neighbours) == 0) next

    overlaps <- sapply(neighbours, function(j) {
      dij <- d[j]
      2 * radius^2 * acos(dij / (2 * radius)) - (dij / 2) * sqrt(4 * radius^2 - dij^2)
    })

    if (any(overlaps > max_overlap)) fail[i] <- TRUE
  }

  df$fail_overlap <- fail
  df
}

flag_intensity <- function(plate, max_intensity_cutoff = 500) {
  plate %>%
    mutate(fail_intensity = max_intensity <= max_intensity_cutoff)
}

flag_perimeter <- function(plate, max_perimeter = 1000) {
  plate %>%
    mutate(fail_perimeter = perimeter >= max_perimeter)
}

apply_filters <- function(plate) {
  plate <- flag_size(plate)
  plate <- flag_edge(plate)
  plate <- flag_intensity(plate)
  plate <- flag_perimeter(plate)
  plate <- plate %>%
    group_by(image_name) %>%
    group_modify(~ flag_overlap_img(.x, radius = radius)) %>%
    ungroup() %>%
    mutate(pass_qc = !fail_size & !fail_edge & !fail_intensity & !fail_perimeter & !fail_overlap)

  n_total <- nrow(plate)
  n_pass  <- sum(plate$pass_qc)

  message(glue(
    "  Filtering summary:\n",
    "    Total:          {n_total}\n",
    "    Fail size:      {sum(plate$fail_size)}\n",
    "    Fail edge:      {sum(plate$fail_edge)}\n",
    "    Fail intensity: {sum(plate$fail_intensity)}\n",
    "    Fail perimeter: {sum(plate$fail_perimeter)}\n",
    "    Fail overlap:   {sum(plate$fail_overlap)}\n",
    "    Pass QC:        {n_pass} ({round(n_pass / n_total * 100, 1)}%)"
  ))

  plate
}

process_plate <- function(filename) {
  experiment_id <- str_extract(basename(filename), "^[^_]+")
  message("Processing: ", experiment_id)

  plate <- read_csv(filename, show_col_types = FALSE) %>%
    mutate(
      Folder_ID   = experiment_id,
      well_row    = as.integer(str_extract(image_name, "(?<=r)\\d+")),
      well_column = as.integer(str_extract(image_name, "(?<=c)\\d+"))
    ) %>%
    left_join(plateID_to_folder_ID, by = "Folder_ID") %>%
    left_join(samplesheet, by = c("well_row", "well_column", "iNDI_Plate_ID")) %>%
    apply_filters()

  message("  → Plate ID: ", unique(plate$iNDI_Plate_ID)[1], 
          " | n = ", nrow(plate), " | pass = ", sum(plate$pass_qc))

  plate
}

all_nuclei <- map(nuc_files, process_plate) %>%
  list_rbind()

dir.create("output_files")
write_parquet(all_nuclei, "output_files/Nuclei_combined_raw.parquet")
```

```{r}
#| label: Filter nuclei 
# that pass QC

good_nuclei <- all_nuclei %>%
  filter(pass_qc == TRUE) %>%
  select(
    Folder_ID,
    image_name,
    label
    )

write_parquet(good_nuclei, "output_files/Nuclei_combined_filtered.parquet")

# Apply to organelle data
message(glue("There are {nrow(df)} rows of organelle data"))

filtered_org <- df %>% 
  semi_join(
    good_nuclei,
    by = c(
      "Measurement_ID" = "Folder_ID",
      "DAPI_filename" = "image_name",
      "Nucleus_ID" = "label"
    )
  )

message(glue("There are {nrow(filtered_org)} rows of organelle data after filtering for nuclei that pass QC"))

write_parquet(filtered_org, "output_files/Organelle_combined_filtered.parquet")
```
